---
title: "Priority effects alter interaction outcomes in a legume-rhizobium mutualism"
author: "Julia Boyle"
date: "14/10/2020"
output: html_document
---
# The set up
```{r}
#Packages required:
library(plyr)
library(tidyverse)
library(car)
library(lme4)
library(lmerTest)
library(cowplot)
library(emmeans)
library(nlme)
library(lsmeans)

#Creating neccessary dataframes:

d<-read.csv('MedicagoJULY.csv')
d$T1_strain= relevel(d$T1_strain, ref="N")
d$T2_strain= relevel(d$T2_strain, ref="N")

d3<-d %>% 
   filter(ExtraNN==0)  #excludes those with biased death prior to any inoculation treatments, most important for mortality tests. The plants excluded by making this data frame are equivalent to 'NN' treatment/control-control treatment. There were 10 extra control-control plants that were allocated to ExtraNN, fewer than 10 survived until the end of the experiment. The inclusion or exclusion of these plants did not change any major results.
d3$T1_strain= relevel(d3$T1_strain, ref="N")
d3$T2_strain= relevel(d3$T2_strain, ref="N")

d2<-d3 %>% 
  mutate(log1pTotal_nod= log1p(Total_nod_num)) #this makes the total nodule distribution normal for later  
d2$T1_strain= relevel(d2$T1_strain, ref="N")
d2$T2_strain= relevel(d2$T2_strain, ref="N")


d4<-d3 %>% 
   filter(Dead1==0)  #excluding plants that died early (before first survey) to help normalize leaf count data
d4$T1_strain= relevel(d4$T1_strain, ref="N")
d4$T2_strain= relevel(d4$T2_strain, ref="N")

d5<-d3 %>% 
  mutate(Leaves_herb= (Leaf4-Leaf4attached)) #Adds a column for number of leaves missing

LeafTime<-read.csv('LeafTime.csv') #mean and SE grouped by treatment for leaf number over time

options(contrasts = c("contr.sum", "contr.poly"))
```

# Key

## Timeline:
- Germinated seeds planted in pots May 23rd, 2019 
- First inoculation was June 6th (14 days after germination)
- Pots transplanted outside June 7th 
- Survey day 1 was June 17th (25 days after germination)
- Second inoculation was June 20th (28 days after germination)
- Survey day 2 was June 25th	(33 days after germination)
- Survey day 3 was July 7th	(45 days after germination)
- Survey day 4/harvest day was July 22nd	(60 days after germination)

## Header meanings
```{r}
head(d3) #What the data sheet looks like
```

- Plant ID is a unique number assigned to each plant
- 'Leaf' columns count the absolute number of leaves created by the plant at each survey timepoint, even if they've been herbivorized later
- 'Dead' columns: 0=alive, 1=dead	
- ExtraNN: These plants were not included in the analyses. Ten extras (NN treatment) that I was going to pull up. In this case, I allocated non-germinated plants (pre-bacterial treatments) into these 10 extras so these should be excluded from death stats due to bias. If the plant survived, the nodule number was still recorded. If any plant died, nodule number was almost always NA because of non-existent roots.
- Pink_nod_num: number of nodules that were pink	
- White nod num: number of nodules that were white	
- Total nod num: total number of nodules on the plant. Marked NA if plant died and had absent roots. Rarely, dead plants still had intact roots with visible nodules.
- Proportion pink: Pink_nod_num/Total_nod_num
- Above ground biomass (g): the above biomass that was still attached to the stem and roots at time of collection. This does not always correspond to number of leaves which was number of ABSOLUTE leaves the plant produced. AGB may be missing/decreased due to herbivory and not treatment related.	

* Treatment: 
  + G= Ensifer meliloti 1022 (Good nitrogen fixing rhizobia)
  + B= Ensifer T173 (Bad nitrogen fixer)
  + N= blank inoculation (No rhizobia) 

# Examining any block effects
```{r}
table(d3$Block, d3$Dead4)#there were 132 plants in block one, 134 in block two. This is because of some seed germination mortality
chisq.test(d3$Block, d3$Dead4)
#Block did not have a significant effect on mortality, above biomass (not shown), or final leaf number (not shown)

d3 %>% 
  group_by(Treatment, Block) %>% 
  count(Treatment) #Treatments pretty equally represented between blocks
```

# Looking at death:models
```{r}
table(d3$Treatment, d3$Dead4)
table(d3$T1_strain, d3$Dead4)
chisq.test(d3$T1_strain, d3$Dead4) #significant
chisq.test(d3$T2_strain, d3$Dead4) #not significant
```

# Looking at total number of nodules and treatment
Models
```{r}
hist(d3$Total_nod_num)
hist(log(d3$Total_nod_num)) #looks normal now

NODNUM <- ddply(d3, c("Treatment"), summarise,
               N    = sum(!is.na(Total_nod_num)),
               mean = mean(Total_nod_num, na.rm=TRUE),
               sd   = sd(Total_nod_num, na.rm=TRUE),
               se   = sd / sqrt(N))
NODNUM

model0<-lm(log1pTotal_nod~T1_strain*T2_strain, data=d2) #using d2 as data because it has the logged values
summary(model0)
Anova(model0, type=3) #significant, getting bacterial treatments increases nodule number compared to getting nothing.
plot(model0)

print(lsmeans(model0.0, list(pairwise ~ T1_strain | T2_strain), type="unlink")) #this gives me predicted means and SE from the model

emm1 = emmeans(model0, specs = ~ T1_strain & T2_strain)
emm1
GBnod<-c(0,0,0,0,0,1,0,0,0)
BGnod<-c(0,0,0,0,0,0,0,1,0)
contrast(emm1, method = list(GBnod - BGnod) ) #planned comparison between T173-1022 and 1022-T173 treatments


T173presentnod<-c(0,1,0,1,1,1,0,1,0)
T173absentnod<-c(1,0,1,0,0,0,1,0,1)
contrast(emm1, method = list(T173presentnod - T173absentnod) ) #do treatments with T173 make more nodules than treatments that do not have it?
```
Boxplot graph
```{r}
levels(d3$T2_strain)
levels(d3$T1_strain) <- c("Control", "T173", "1022")
levels(d3$T2_strain) <- c("Control", "T173", "1022")
d3$Treatment<-factor(d3$Treatment, levels=c("NN", "NB", "NG", "BN", "BB", "BG", "GB", "GN", "GG"))
nodnumbox2<-ggplot(data=d3, aes(x=T2_strain, y=Total_nod_num, fill=Treatment))+
  geom_boxplot()+
  facet_grid(.~ T1_strain, space = 'free_x', scales = 'free_x', switch = 'x') +
  labs(x = "First inoculation")+
  theme_cowplot()
nodnumbox2labelled<-nodnumbox2+ xlab('Treatment') + ylab('Nodules (no.)') +theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+scale_fill_manual(values=c("white", "white", "white","white", "white", "grey","grey", "white", "white"))+ theme(legend.position="none") 
nodnumbox2labelled
ggsave2("nodulenumbox2.png", nodnumbox2labelled, dpi=600, width=7, height=4)
```

# Looking at proportion of nodules that are nitrogen fixing
Old Models
```{r}
NODSIMPLE <- ddply(d3, c("Treatment"), summarise,
               N    = sum(!is.na(Proportion_pink)),
               mean = mean(Proportion_pink, na.rm=TRUE),
               sd   = sd(Proportion_pink, na.rm=TRUE),
               se   = sd / sqrt(N))
NODSIMPLE

hist(d3$White_nod_num)
hist(log1p(d3$White_nod_num))
hist(d3$Pink_nod_num)
hist(log1p(d3$Pink_nod_num))

model11<-manova(cbind(log1p(White_nod_num), log1p(Pink_nod_num)) ~ T1_strain*T2_strain, data = d3)
summary(model11)#MANOVA F, df, p come from here
summary.aov(model11)
model11$coefficients
str(model11)

model11.2<-lm(log1p(Pink_nod_num)~T1_strain*T2_strain, data=d) 
summary(model11.2)#Effective nodule estimates and error come from here
Anova(model11.2, type=3)#Effective nodule F, df, and p come from here
model11.2$coefficients

model11.3<-lm(log1p(White_nod_num)~T1_strain*T2_strain, data=d) 
summary(model11.3) #Ineffective nodule estimates and error come from here
Anova(model11.3, type=3) #Ineffective nodule F, df, and p come from here
model11.3$coefficients
```
New model for nodule occupancy
```{r}
model11.review3<-glm(cbind(Pink_nod_num, Total_nod_num)~T1_strain*T2_strain, family="quasibinomial", weights=Total_nod_num, data=d3)
summary(model11.review3)
Anova(model11.review3, type=3)#all significant
plot(model11.review3)

emm2.3 = emmeans(model11.review3, specs = ~ T1_strain & T2_strain)
emm2.3
GBPinknum<-c(0,0,0,0,0,1,0,0,0)
BGPinknum<-c(0,0,0,0,0,0,0,1,0)
contrast(emm2.3, method = list(BGPinknum - GBPinknum)) #planned comparison between 1022-T173 and T173-1022 treatments for number of pink nodules. This is the one now displayed on figure

lsmeans(model11.review3, list(pairwise ~ T2_strain | T1_strain), type="unlink") #back-transformed
lsmeans(model11.review3, list(pairwise ~ T2_strain | T1_strain))
```
Boxplot graph
```{r}
levels(d3$T2_strain)
levels(d3$T1_strain) <- c("Control", "T173", "1022")
levels(d3$T2_strain) <- c("Control", "T173", "1022")
d3$Treatment<-factor(d3$Treatment, levels=c("NN", "NB", "NG", "BN", "BB", "BG", "GB", "GN", "GG"))
proportionbox2<-ggplot(data=d3, aes(x=T2_strain, y=Proportion_pink, fill=Treatment))+
  geom_boxplot()+
  facet_grid(.~ T1_strain, space = 'free_x', scales = 'free_x', switch = 'x') +
  labs(x = "First inoculation")+
  theme_cowplot()
proportionbox2labelled<-proportionbox2+ xlab('Treatment') + ylab('Proportion of nitrogen-fixing nodules') +theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+scale_fill_manual(values=c("white", "white", "white","white", "white", "grey","grey", "white", "white"))+ theme(legend.position="none") 
proportionbox2labelled
ggsave2("proportionbox2.png", proportionbox2labelled, dpi=600, width=7, height=5)
```
# Above biomass
Model
```{r}
BiomassTreat <- ddply(d3, c("Treatment"), summarise,
               N    = sum(!is.na(Above_biomass)),
               mean = mean(Above_biomass, na.rm=TRUE),
               sd   = sd(Above_biomass, na.rm=TRUE),
               se   = sd / sqrt(N))
BiomassTreat

model5.3<- glmer(Above_biomass~T1_strain*T2_strain + (1 | Block), family=Gamma, data=d3)  
summary(model5.3) 
Anova(model5.3, type=3) #t1 strain sig
plot(model5.3)

lsmeans(model5.3, list(pairwise ~ T2_strain | T1_strain)) #this gives lsmean in the transformed way, 'inverse'
lsmeans(model5.3, ~ T2_strain | T1_strain, type = "unlink") #this gives me untransformed 

emm3 = emmeans(model5.3, specs = ~ T1_strain & T2_strain)
emm3
GBagb<-c(0,0,0,0,0,1,0,0,0)
BGagb<-c(0,0,0,0,0,0,0,1,0)
contrast(emm3, method = list(GBagb - BGagb)) #planned comparison between 1022-T173 and T173-1022 treatments
```
Boxplot graph
```{r}
levels(d3$T2_strain)
levels(d3$T1_strain) <- c("Control", "T173", "1022")
levels(d3$T2_strain) <- c("Control", "T173", "1022")
d3$Treatment<-factor(d3$Treatment, levels=c("NN", "NB", "NG", "BN", "BB", "BG", "GB", "GN", "GG"))
agbbox<- d3 %>% 
  filter(Above_biomass<0.05) %>% 
  ggplot(aes(x=T2_strain, y=Above_biomass, fill=Treatment))+
  geom_boxplot()+
  facet_grid(.~ T1_strain, space = 'free_x', scales = 'free_x', switch = 'x') +
  labs(x = "First inoculation")+
  theme_cowplot()
agbboxlabelled<-agbbox+ xlab('Treatment') + ylab('Aboveground biomass (g)') +theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+scale_fill_manual(values=c("white", "white", "white","white", "white", "grey","grey", "white", "white"))+ theme(legend.position="none") 
agbboxlabelled
ggsave2("agbbox.png", agbboxlabelled, dpi=600, width=7, height=5)
```

# Supplementary
## Nodule colour on biomass: plots
Number of nodules
```{r}
#How do white nodules predict biomass
WHITENODBIO<-ggplot(data=d3, aes(x=White_nod_num, y=Above_biomass))+
  geom_point()+
  geom_smooth(method=lm)+
  theme_cowplot()#Negative slope, more white nodules decreases biomass
WHITENODBIOFINAL<-WHITENODBIO+ xlab('Number of ineffective nodules (no.)') + ylab('Aboveground biomass (g)')
WHITENODBIOFINAL
#ggsave2("WhiteNoduleBiomass.png",WHITENODBIOFINAL, dpi=600)
modelsup1<- lm(Above_biomass~White_nod_num, data=d3) 
summary(modelsup1)
Anova(modelsup1, type=3)

#Let's see how well number of nitrogen fixing nodules predicts biomass
PINKNODBIO<-ggplot(data=d3, aes(x=Pink_nod_num, y=Above_biomass))+
  geom_point()+
  geom_smooth(method=lm)+
  theme_cowplot()#Positive slope, more pink nodules increases biomass
PINKNODBIOFINAL<-PINKNODBIO+ xlab('Number of nitrogen-fixing nodules (no.)') + ylab('Aboveground biomass (g)')
PINKNODBIOFINAL
modelsup2<- lm(Above_biomass~Pink_nod_num, data=d3) 
summary(modelsup2)
Anova(modelsup2, type=3)

ggsave2("PinkNoduleBiomass.png",PINKNODBIOFINAL, dpi=600)
ggsave2("WhiteNoduleBiomass.png",WHITENODBIOFINAL, dpi=600)
```
Proportion of nodules
```{r}
ProportionAGB<-ggplot(data=d3, aes(x=Proportion_pink, y=Above_biomass))+
  geom_point()+
  geom_smooth(method=lm)+
  theme_cowplot()
ProportionAGBlabelled<-ProportionAGB+ xlab('Proportion of nitrogen-fixing nodules') + ylab('Aboveground biomass (g)')
ProportionAGBlabelled
ggsave2("ProportionAGBsup.png", ProportionAGBlabelled, dpi=600)

modelsup3<-glm(Above_biomass~cbind(Pink_nod_num, Total_nod_num), family=Gamma, weights=Total_nod_num, data=d3)
summary(modelsup3)
Anova(modelsup3, type=3)

```
## Model and plot for first strain and nodule number effect on biomass
```{r}
model7.1<-glmer(Above_biomass~Total_nod_num*T1_strain + (1 | Block), family=Gamma, data=d3) 
summary(model7.1) 
Anova(model7.1, type=3) #T1_strain marginally significant, total nod num sig
plot(model7.1) 

#How this looks visually:
BiomassNodNum<-ggplot(data=d3, aes(x=Total_nod_num, y=Above_biomass, colour=T1_strain))+
  geom_point()+
  geom_smooth(method=lm)+
  scale_color_manual(values=c('#3182BD','#DE2D26', '#2CA25F'), labels=c("Control", "T173", "1022"))+
  theme_cowplot()
BiomassNodNumFinal<-BiomassNodNum +xlab('Total number of nodules (no.)') + ylab('Aboveground biomass (g)') +labs(colour='First inoculation') 
BiomassNodNumFinal 
ggsave2("BiomassNoduleNumber.png", BiomassNodNumFinal, dpi=600)
```
## Leaf data and Treatment: Plot and model
```{r}
LEAFTREAT1 <- ddply(d3, c("Treatment"), summarise,
               N    = sum(!is.na(Leaf1)),
               mean = mean(Leaf1, na.rm=TRUE),
               sd   = sd(Leaf1, na.rm=TRUE),
               se   = sd / sqrt(N))
LEAFTREAT1
LEAFTREAT2 <- ddply(d3, c("Treatment"), summarise,
               N    = sum(!is.na(Leaf2)),
               mean = mean(Leaf2, na.rm=TRUE),
               sd   = sd(Leaf2, na.rm=TRUE),
               se   = sd / sqrt(N))
LEAFTREAT2
LEAFTREAT3 <- ddply(d3, c("Treatment"), summarise,
               N    = sum(!is.na(Leaf3)),
               mean = mean(Leaf3, na.rm=TRUE),
               sd   = sd(Leaf3, na.rm=TRUE),
               se   = sd / sqrt(N))
LEAFTREAT3
LEAFTREAT4 <- ddply(d3, c("Treatment"), summarise,
               N    = sum(!is.na(Leaf4)),
               mean = mean(Leaf4, na.rm=TRUE),
               sd   = sd(Leaf4, na.rm=TRUE),
               se   = sd / sqrt(N))
LEAFTREAT4 #all these summarized dataframes went into making the LeafTime.csv file

LeafTimePlot<-LeafTime %>% 
  group_by(Treatment, Time) %>% 
  ggplot(aes(x=Time, y=MeanLeaf, colour=Treatment))+
  geom_point(size=3, position=position_dodge(width=4))+
  geom_errorbar(aes(ymin=MeanLeaf-SELeaf, ymax=MeanLeaf+SELeaf), width=.1, position=position_dodge(width=4))+
  geom_line(position=position_dodge(width=4))+
  scale_color_manual(name = 'Treatment', values=c('#66C2A4','#2CA25F', '#006D2C', '#6BAED6', '#3182BD', '#08519C', '#A50F15', '#DE2D26', '#FB6A4A'))+
  theme_cowplot()
LeafTimePlotFinal<-LeafTimePlot+ xlab('Days since germination')+ylab('Leaves (no.)')+ geom_segment(aes(x=14, xend=14, y=0, yend=5), linetype=2) + annotate("text", x = 10, y = 5.2, label = "First inoculation", size=4)+ geom_segment(aes(x=28, xend=28, y=0, yend=5), linetype=2) + annotate("text", x = 33, y = 5.2, label = "Second inoculation", size=4)
LeafTimePlotFinal
#ggsave2("LeafTimePlotFinal.png",LeafTimePlotFinal, dpi=600)

model3<- lmer(Leaf4~T1_strain*T2_strain + (1 | Block), data=d4)#Using the dataframe that excludes plants that were dead after first survey
summary(model3)
Anova(model3, type=3) #only interaction stays sig
plot(model3)#normal
```
## Herbivory/Senescent leaves model
```{r}
model10<- manova(cbind(Leaf4, Leaf4attached) ~ T1_strain*T2_strain, data = d3) #T1_strain is significant
summary(model10)
summary.aov(model10)
```